Evaluation of +S2 sea ice forecast and some other better title

### 2025-05-07 analysis/paper/access-ice.tex
Revision

Antarctic sea ice is in a remote location, but the potential impacts of
its variability extends far into the lower latitudes, affecting the
circulation of both the ocean and the atmosphere, ocean carbon uptake
and biological processes. It is likely that a reduction in sea ice
extent would reduce the meridional temperature gradient which, in turn,
would reduce the strength and location of the eddy-driven jet (and the
opposite circulation changes with an increase in sea ice), which is a
crucial control on extratropical weather. In the northern hemisphere
(NH), it has been hypothesised that the reduction in Arctic sea ice has
led to a weaker, wavier jet, increasing the frequency of extreme events
\citep{barnes2015}. There is some evidence of this effect in the
southern hemisphere (SH). Models show a weakened vortex in the middle
atmosphere, increased vertical wave flux and greater zonal wave 1
amplitude in austral spring following a low sea ice maximum. But models
do not fully agree on the exact phase of the zonal wave response, and
observations do not show a clear signal \citep{rea2024}. The jet
response to sea ice loss is also very model dependent, particularly
sensitive to its climatological location, and relatively small compared
with the direct effect of CO2 increase and sea surface temperature (SST)
warming \citep{ayres2019}. On climate scales, it is thought that the
temperature signal of a reduction in Antarctic sea ice mirrors a ``mini
global warming'', and experiments also predict a weakening and
equatorward shift of the SH eddy-driven jet
\citep{raphael2011, england2018, england2020, ayres2022}.

Correctly modelling Antarctic sea is not only necessary for process
understanding and climate projections to inform adaptation strategies
but also for accurate seasonal to sub-seasonal forecasts, which are
crucial for operations in and around the Antarctic continent such as
scientific missions, fisheries and tourism \citep[
\citet{wagner2020}]{desilva2020}. However, producing accurate Antarctic
sea ice forecasts has been challenging due to model biases, inherent
large variability and complexity, and it has lagged behind Arctic sea
ice forecasts \citep{zampieri2019}. Dynamical seasonal forecasts of
summer Antarctic sea ice have been shown to perform worse than
relatively simple statistical methods \citep{massonnet2023}, which also
underscores the need for better understanding of sea ice dynamics.

@@

The ACCESS-S2 hindcast used in this study runs for the period
1981--2018. Each forecast consists of 9 ensemble members built from
three consecutive 3-member forecasts initialised at the first of every
month and the two previous days and run from 279 days. The ACCESS-S1
hindcast is built in the same manner and running for the period
1990--2012.

@@

The European Organisation for the Exploitation of Meteorological
Satellites (EUMETSAT) Ocean and Sea Ice Satellite Application Facility
(OSI SAF) \citep{OSISAF} is another satellite-derived sea ice
concentration product. It is based on mostly the same sensors as the
NOAA CDR but computed independently using different algorithms. Figures
prepared with this dataset are provided in the supplementary material.

@@

We compute RMSE as the square root of the average squared differences
between forecasted and observed sea ice concentration anomalies. We
compute a pan-Antarctic RMSE by averaging over the whole NOAA/NSIDC
CDRV4 southern hemisphere domain, and also a zonally-varying RMSE by
averaging over twenty-four 15° longitude slices around Antarctica.

@@

\caption{\label{fig-hindcast-extent}Median sea ice extent for al
hindcasts initialised the first of the month for ACCESS-S2 and ACCESS-S1
in colours representing the start month with the median sea ice extent
of NOAA/NSIDC CDRV4 in the left column and the median difference in the
right column. The NOAA/NSIDC CDRV4 climatology is computed in the period
corresponding to each hindcast. Circles represent the initial conditions
at the first of every month and triangles represent the median value at
the first of every month forecasted with the largest possible lead time.
Colours indicate the initialisation month of the forecast.}

@@

Both models' equilibrium state (triangles) show a negative bias of sea
ice extent, particularly in the late-autumn and winter months. This is
due primarily to faster melt during a longer melt season between January
and March and slower growth during March and April. This is then
partially balanced with faster growth between May and July
(Fig.~\ref{fig-mean-growth}). Many sea ice models exhibit this
systematic underestimation during the sea ice minimum and early freezing
season \citep{massonnet2023} and could indicate problems in the
representation of thermodynamics in the model \citep{zampieri2019}. It
is not surprising that both forecasting systems converge to a similar
climatology because they share the same model formulation.

ACCESS-S2 initial conditions (circles) also have a negative bias,
especially in the late summer-early autumn, while ACCESS-S1 initial
conditions are very close to observations. The faster growth between May
and July reduces the negative bias in ACCESS-S2 for forecasts
initialised from April to June, but it adds a positive bias in
ACCESS-S1.

@@

\caption{\label{fig-mean-growth}Median daily sea ice extent growth of
ACCESS-S1 and ACCESS-S2 hindcasts and observations. Values are smoothed
with a 11-day running mean.}

@@

\caption{\label{fig-bias}Mean sea ice concentrations 1-month lag monthly
mean ACCESS-S2 forecast bias compared with NSIDC.}

@@

Figure~\ref{fig-bias} shows the difference of monthly mean sea ice
concentrations between NOAA/NSIDC CDRV4 and ACCESS-S2 hindcasts at the
shortest lag. From October to May, the model underestimates sea ice
concentrations in most regions except for the inner Weddell Sea in April
and May, where sea ice concentrations saturate to 1 both in the
observations and forecasts. In winter, the differences are mostly on the
sea ice edge with light positive bias in the African sector of East
Antarctica and negative bias around the Indian Ocean sector which
partially compensate, resulting in the near-zero extent bias seen in
those months (Fig.~\ref{fig-hindcast-extent}).

@@

\caption{\label{fig-extent-anom}Monthly mean sea ice extent anomalies
for ACCESS-S1 and ACCESS-S2 (black) and NOAA/NSIDC CDRV4 (blue)
forecasted at selected lead times. The RMSE during the overlapping
period (1990--2013) is shown on the top left.}

@@

Figure~\ref{fig-extent-anom} shows monthly sea ice extent anomalies
forecasted at selected lead times. Compared with ACCESS-S1, ACCESS-S2
anomaly forecast is relatively poor even in the first month, which stays
relatively skilful even at lag 3. ACCESS-S2 shows much bigger
variability than observations, with dramatic lows between 1995 and 2007
and highs between 2007 and 2015.

@@

\caption{\label{fig-rmse}Median RMSE of sea ice concentration anomalies
as a function of forecast lag for all forecast initialised on the first
of each month compared with a reference forecast of persistence of
anomalies.}

@@

monthly rmse/corelation at gridpoints per lead time.

Figure~\ref{fig-rmse} shows the median RMSE of sea ice concentration
anomalies for ACCESS-S2 and ACCESS-S1 hindcasts compared with a
benchmark of persistence and climatology. Due to errors in the initial
conditions, it is expected that a persistence forecast would be better
than the model forecast at very short lead times, but that the
persistence forecast errors would grow faster and eventually surpass the
model forecast, after when the model is statistically useful. Here the
persistence errors are almost always lower than the ACCESS-S2 forecast,
indicating that the model does not have skill at any lead time and in
any month. The only exception is around February, where the model has
lower RMSE than the persistence forecast at virtually every lag.

@@

\caption{\label{fig-rmse_lon}Median difference between RMSE of ACCESS-S2
forecasts and persistence forecast computed on
\texttt{r\ uniqueN(rmse\_lon\_mean\$lon)} meridional slices
\texttt{r\ diff(unique(rmse\_lon\_mean\$lon)){[}1{]}}º wide.}

@@

The skill shown by ACCESS-S2 at February-March forecasts is evidenced by
a band of negative values across almost all longitudes. The only region
where February-March sea ice is not well forecasted is east of the
Antarctic Peninsula, which is the only region with consistent summer sea
ice. This suggests that the positive skill around the summer minimum
comes from the model correctly forecasting no ice where no ice is ever
found. However, there are other regions of skilful forecasts.

Forecasts initialised on January 1st (Fig.~\ref{fig-rmse_lon} panel a)
show RMSE values below persistence at lags larger than 180 days
(corresponding to July through September) in the Ross and Weddell Sea.
These two regions are also relatively well forecasted from other months.
Notably, May forecasts of Weddell sea ice concentration anomalies
(Fig.~\ref{fig-rmse_lon} panel e) are skilful after about 20 days and
remain so until December. The Weddell Sea is also the region with the
maximum error --around June regardless of initialisation date.

@@

\caption{\label{fig-iiee}Median and 95\% coverage of Integrated Ice Edge
Error as a function of forecast lag for all forecast initialised on the
first of each month for ACCESS-S1 and ACCESS-S2 hindcasts. For each
month, the number in parenthesis indicates the minimum lag at which the
the mean error of each model is not statistically significant at a 1\%
level using a two-sisded t-test.}

@@

Also evident in Figure~\ref{fig-iiee} is the difference in the error
spread at short lags between ACCESS-S2 and ACCESS-S1. In all months
ACCESS-S1 has much narrower error spread at the first day of the
forecasts. The error in the initial conditions is not only small, but
also fairly constant. This spread then grows towards a climatological
spread as errors accumulate differently in different forecasts. For
ACCESS-S2, this is true only between July and October, approximately.
For all other months, the error spread is more or less stable throughout
the forecast window, indicating that not only the initial error is high,
but it is not constant.

@@

\caption{\label{fig-iiee-variance}Mean spread of IIEE at different lags
for different models.}

@@

Sea ice forecasts from the ACCESS-S2 model show a significant low extent
bias, particularly during late summer and early autumn. This bias is
attributed to a faster and longer melt season between January and March,
and slower growth between March and April. This underestimation during
the minimum and early freezing season is a common issue in many
seasonal-to-subseasonal (S2S) systems, suggesting potential problems
either with the model's thermodynamic representation or with short wave
radiation forcing, as shown in other climate models. Even though
ACCESS-S2 shares the same model formulation with ACCESS-S1, the latter
does not suffer from this bias. This is due to the assimilation of sea
ice concentrations into the initial conditions, which successful
corrects for the negative bias in the model. Our analysis suggests that
the data assimilation system in ACCESS-S2 is only effectively
influencing sea ice initial conditions from June to October, while the
rest of the year, the sea ice component runs virtually free, reverting
to its biased equilibrium state.

>>>>

Antarctic sea ice covers the polar ocean, yet the potential impacts of
its variability extend far into the mid and lower latitudes, affecting
the circulation of both the ocean and the atmosphere, ocean carbon
uptake and biological processes \citep[\citet{england2020},
\citet{swadling2023}]{fogwill2020}. In coupled climate model
experiments, a reduction in sea ice extent has been shown to reduce the
meridional temperature gradient, in turn reducing the eddy-driven jet
strength and shifting it equatorward \citetext{\citealp[
\citet{england2020}]{england2018}; \citealp{ayres2019}; \citealp{ayres2022}}
and vice versa for an increase in sea ice
\citep{raphael2003a, raphael2011}, which is a crucial control on
extratropical weather. In the Northern Hemisphere, it has been
hypothesised that the reduction in Arctic sea ice has led to a weaker,
wavier jet, increasing the frequency of extreme events
\citep{barnes2015, smith2022b}. There is some evidence of this effect in
the Southern Hemisphere. For example, model simulations show a weakened
vortex in the middle atmosphere, increased vertical wave flux and
greater zonal wave 1 amplitude in austral spring following lower than
average sea ice extent at in early winter
\citep[\citet{song2025}]{rea2024}. But models different models disagree
on the exact phase of the zonal wave response, and observations do not
show a clear signal \citep{rea2024}. The magnitude of the eddy-driven
jet response to sea ice loss has been shown to be model dependent, and
is sensitive to its climatological location \citep{ayres2019}. Further,
sea ice induced changes in the jet have been found to be relatively
small compared with the jet-response to the direct effect of CO2
increase and global sea surface temperature (SST) warming
\citep{ayres2022}.

Correctly modelling Antarctic sea is not only necessary for process
understanding and climate projections to inform adaptation strategies
but also for accurate seasonal to sub-seasonal forecasts, which are
crucial for operations in and around the Antarctic continent such as
scientific missions, fisheries and tourism \citep[
\citet{wagner2020}]{desilva2020}. However, producing accurate Antarctic
sea ice forecasts has been challenging due to model biases, inherent
large variability and complexity, and it has lagged behind Arctic sea
ice forecasts \citep{zampieri2019}. Dynamical seasonal forecasts of
summer Antarctic sea ice have been shown to perform worse than
relatively simple statistical methods \citep{massonnet2023}, which also
underscores the need for better understanding of sea ice dynamics, and
drivers.

@@

The ACCESS-S2 hindcast used in this study runs for the period
1981--2018. Each forecast consists of 9 ensemble members built from
three consecutive 3-member forecasts initialised at the first of every
month and the two previous days and run from 279 days. The ACCESS-S1
hindcast is built in the same manner (albeit with 217-day runs) for the
period 1990--2012.

@@

The European Organisation for the Exploitation of Meteorological
Satellites (EUMETSAT) Ocean and Sea Ice Satellite Application Facility
(OSI SAF) \citep{OSISAF} is another satellite-derived sea ice
concentration product. It is based on mostly the same sensors as the
NOAA CDR but computed independently using different algorithms. Figures
prepared with this dataset are provided in the supplementary material
and don't differ significantly from the ones prepared using CDR.

@@

We compute RMSE as the square root of the average squared differences
between forecasted and observed sea ice concentration anomalies. We
compute a pan-Antarctic RMSE by averaging over the whole NOAA/NSIDC
CDRV4 southern hemisphere domain, and also a zonally-varying RMSE
computed over twenty-four 15° longitude slices around Antarctica.

@@

\caption{\label{fig-hindcast-extent}Pan-antarctic median sea ice extent
for al hindcasts initialised the first of the month for ACCESS-S2 and
ACCESS-S1 in colours representing the start month with the median sea
ice extent of NOAA/NSIDC CDRV4 in the left column and the median
difference in the right column. The NOAA/NSIDC CDRV4 climatology is
computed in the period corresponding to each hindcast. Circles represent
the initial conditions at the first of every month and triangles
represent the median value at the first of every month forecasted with
the largest possible lead time. Colours indicate the initialisation
month of the forecast.}

@@

ACCESS-S2 initial conditions (circles) show an overall negative bias,
especially in the late summer-early autumn, while ACCESS-S1 initial
conditions are very close to observations. Both models' equilibrium
state (triangles) show a negative bias of sea ice extent, particularly
in the late-autumn and winter months. This is due primarily to faster
melt during a longer melt season between January and March and slower
growth during March and April. This is then partially balanced with
faster growth between May and July (Fig.~\ref{fig-mean-growth}). Many
sea ice models exhibit this systematic underestimation during the sea
ice minimum and early freezing season \citep{massonnet2023} and could
indicate problems in the representation of thermodynamics in the model
\citep{zampieri2019}. It is not surprising that both forecasting systems
converge to a similar climatology because they share the same model
formulation.

@@

\caption{\label{fig-mean-growth}Median daily sea ice extent growth of
ACCESS-S1 and ACCESS-S2 hindcasts and observations computed as the
median daily differences in sea ice extent for each date and all lead
times. Values are smoothed with a 11-day running mean.}

@@

\caption{\label{fig-bias}Mean difffernece between monthly sea ice
concentrations at 0-month lead time ACCESS-S2 forecast and
observations.}

@@

Figure~\ref{fig-bias} shows the difference of monthly mean sea ice
concentrations between NOAA/NSIDC CDRV4 and ACCESS-S2 hindcasts at the
shortest lead time From October to May, the model underestimates sea ice
concentrations in most regions except for the inner Weddell Sea in April
and May, where sea ice concentrations saturate to 1 both in the
observations and forecasts. In winter, the differences are mostly on the
sea ice edge with light positive bias in the African sector of East
Antarctica and negative bias around the Indian Ocean sector which
partially compensate, resulting in the near-zero extent bias seen in
those months (Fig.~\ref{fig-hindcast-extent}).

@@

\caption{\label{fig-extent-anom}Monthly mean sea ice extent anomalies
for ACCESS-S1 and ACCESS-S2 and observations (black) forecasted at
selected lead times. The RMSE during the overlapping period (1990--2013)
is shown on the top left.}

@@

Figure~\ref{fig-extent-anom} shows monthly sea ice extent anomalies
forecasted at selected lead times. Compared with ACCESS-S1, ACCESS-S2
anomaly forecast is relatively poor even in the first month, which stays
relatively skilful even at a lead time of 3 months. ACCESS-S2 shows much
bigger variability than observations, with dramatic lows between 1995
and 2007 and highs between 2007 and 2015.

@@

\caption{\label{fig-rmse}Median RMSE of sea ice concentration anomalies
as a function of forecast lead time for all forecast initialised on the
first of each month compared with a reference forecast of persistence of
anomalies (black) and climatology (gray).}

@@

Figure~\ref{fig-rmse} shows the median RMSE of sea ice concentration
anomalies for ACCESS-S2 and ACCESS-S1 hindcasts compared with a
benchmark of persistence and climatology. Due to errors in the initial
conditions, it is expected that a persistence forecast would be better
than the model forecast at very short lead times, but that the
persistence forecast errors would grow faster and eventually surpass the
model forecast, after when the model is statistically useful. Here the
persistence errors are almost always lower than the ACCESS-S2 forecast,
indicating that the model does not have skill at any lead time and in
any month. The only exception is around February, where the model has
lower RMSE than the persistence forecast at virtually every lead time.

@@

\caption{\label{fig-rmse_lon}Median difference between RMSE of ACCESS-S2
forecasts and persistence forecast computed on
\texttt{r\ uniqueN(rmse\_lon\_mean\$lon)} meridional slices
\texttt{r\ diff(unique(rmse\_lon\_mean\$lon)){[}1{]}}º wide.
Antarctica's coastline is shown at the bottom of each panel for
reference.}

@@

The skill shown by ACCESS-S2 at February-March forecasts is observed

by a band of negative values across almost all longitudes. The only
region where February-March sea ice is not well forecasted is east of
the Antarctic Peninsula, which is the only region with consistent summer
sea ice. This suggests that the positive skill around the summer minimum
comes from the model correctly forecasting no ice where no ice is ever
found. However, there are other regions of skilful forecasts.

Forecasts initialised on January 1st (Fig.~\ref{fig-rmse_lon} panel a)
show RMSE values below persistence at lead times larger than 180 days
(corresponding to July through September) in the Ross and Weddell Sea.
These two regions are also relatively well forecasted from other months.
Notably, May forecasts of Weddell sea ice concentration anomalies
(Fig.~\ref{fig-rmse_lon} panel e) are skilful after about 20 days and
remain so until December. The Weddell Sea is also the region with the
maximum error --around June regardless of initialisation date.

@@

\caption{\label{fig-iiee}Median and 95\% coverage of Integrated Ice Edge
Error as a function of forecast lead time for all forecast initialised
on the first of each month for ACCESS-S1 and ACCESS-S2 hindcasts. For
each month, the number in parenthesis indicates the minimum lead time at
which the the mean error of each model is not statistically significant
at a 1\% level using a two-sisded t-test.}

@@

Also evident in Figure~\ref{fig-iiee} is the difference in the error
spread at short lead times between ACCESS-S2 and ACCESS-S1. In all
months ACCESS-S1 has much narrower error spread at the first day of the
forecasts. The error in the initial conditions is not only small, but
also fairly constant. This spread then grows towards a climatological
spread as errors accumulate differently in different forecasts. For
ACCESS-S2, this is true only between July and October. For all other
months, the error spread is more or less stable throughout the forecast
window, indicating that not only the initial error is high, but it is
not constant.

@@

\caption{\label{fig-iiee-variance}Mean spread of IIEE at different lead
times for different models.}

@@

Sea ice forecasts from the ACCESS-S2 model show a significant low extent
bias, particularly during late summer and early autumn. This bias is
attributed to a faster and longer melt season between January and March,
and slower growth between March and April. This underestimation during
the minimum and early freezing season is a common issue in many
seasonal-to-subseasonal (S2S) systems, suggesting potential problems
either with the model's thermodynamic representation or with short wave
radiation forcing, as shown in other climate models
\citep{zampieri2019, roach2020}. Even though ACCESS-S2 shares the same
model formulation with ACCESS-S1, the latter does not suffer from this
bias. This is due to the assimilation of sea ice concentrations into the
initial conditions, which successful corrects for the negative bias in
the model. Our analysis suggests that the data assimilation system in
ACCESS-S2 is only effectively influencing sea ice initial conditions
from June to October, while the rest of the year, the sea ice component
runs virtually free, reverting to its biased equilibrium state.
